Covariates analysis of MSBB reprocessed RNASeq data

### 2016-05-31 code/Rmd/MSSM_RNASeq_cdr_Reprocessed.Rmd
Added markdown to process reprocessed rnaseq data from MSSM batch 1,2 and 3

---
title: "Covariates analysis of MSBB RNASeq data"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)
library(githubr)

@@

knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_cdr.Rmd",
                                 parentId ="syn5570248",
                                 entityName = 'MSBB RNASeq (based on CDR)')
```

@@

library(ggplot2)
library(reshape2)
library(limma)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(xlsx)

@@

cl = makeCluster(12)
registerDoParallel(cl)

@@

thisFileName <- 'MSSM_RNASeq_cdr.Rmd'

@@

Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn5898488'
ALL_USED_IDs <- COUNT_ID
COUNT_OBJ <- synGet(COUNT_ID)
COUNT <- fread(COUNT_OBJ@filePath, data.table = F)
rownames(COUNT) = COUNT$Geneid

# Get technical metadata
METADATA_ID <- 'syn5898489'
METADATA_OBJ <- synGet(METADATA_ID)
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA <- read.table(METADATA_OBJ@filePath,sep='\t',header=T, row.names=1)

# Get clinical metadata
METADATA_TECHNICAL_ID <- 'syn5475828'
METADATA_TECHNICAL_OBJ <- synGet(METADATA_TECHNICAL_ID)
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_TECHNICAL_ID
METADATA_TECHNICAL <- read.xlsx(METADATA_TECHNICAL_OBJ@filePath, sheetIndex = 1)

# Manually fix 4 sample IDs
METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1132), "BM36.Fnl"] = 397
METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1117), "BM22.Fnl"] = 144
METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1320), "BM22.Fnl"] = 159
METADATA_TECHNICAL[which(METADATA_TECHNICAL$BB == 1475), "BM22.Fnl"] = 186

METADATA = left_join(METADATA %>%
                       dplyr::mutate(ID = gsub('hB_RNA_','',NewBarcode),
                                     ID = as.numeric(ID)),
                     METADATA_TECHNICAL %>%
                       dplyr::select(BB, BM.44.Barcode, BM36.Fnl, BM22.Fnl, BM10.Fnl, PMI, RACE, 
                                     AOD, CDR, SEX, NP.1, PlaqueMean, bbscore) %>%
                       dplyr::rename(BM_44 = BM.44.Barcode,
                                     BM_10 = BM10.Fnl,
                                     BM_22 = BM22.Fnl,
                                     BM_36 = BM36.Fnl) %>%
                       tidyr::gather(BrainRegion, ID, -BB, -PMI, -RACE, 
                                     -AOD, -CDR, -SEX, -NP.1, -PlaqueMean, -bbscore) %>%
                       filter(!is.na(ID)) %>%
                       dplyr::mutate(ID = as.numeric(ID)))
rownames(METADATA) = METADATA$NewBarcode

# Filter data
ind <- intersect(rownames(METADATA), colnames(COUNT))
ind <- setdiff(ind, 'hB_RNA_10622')
METADATA <- METADATA[ind,]
COUNT <- COUNT[, ind]
```
### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering, cache=TRUE}
# Max out age at 90
METADATA$AOD = gsub('\\+','',as.character(METADATA$AOD))

@@

# Select covariates of interest
FactorCovariates <- c('RACE', 'SEX', 'BrainRegion.CDR', 'Batch', 'BB')
ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2', 'PlaqueMean')

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$NewBarcode

@@

```{r dataExplore, fig.width = 10, fig.height = 10}
# RIN
p = list()
p[[1]] = ggplot(METADATA, aes(x = CDR, y = RIN, fill = BrainRegion)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')

# AgeAtDeath
p[[2]] = ggplot(METADATA, aes(x = CDR, y = as.numeric(AOD), fill = BrainRegion)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + theme(legend.position = 'top')

# Mean Plaque
p[[3]] = ggplot(METADATA, aes(x = CDR, y = as.numeric(PlaqueMean), fill = BrainRegion)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PlaqueMean') + theme(legend.position = 'top')

# NP.1
p[[4]] = ggplot(METADATA, aes(x = CDR, y = as.numeric(NP.1), fill = BrainRegion)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('NP.1') + theme(legend.position = 'top')

@@

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,FactorCovariates]))
dev.off()
```
### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES,'NULL design(voom-normalized)', isKeyPlot=TRUE, MIN_PVE_PCT_PC = 1)
```
Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = rownames(RESIDUAL.GENE_EXPRESSION),
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id"),
                                      Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
VOOM.ADJUSTED.GENE_EXPRESSION$E <- left_join(rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, "ensembl_gene_id"),
                                             Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = as.data.frame(summary(COVARIATES$BrainRegion.CDR))
colnames(tmp) = 'No. of samples'
kable(tmp)
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=18}
# Fit contrast
contrast = makeContrasts(contrasts = sapply(unique(METADATA$BrainRegion), function(x){
  apply(combn(paste0('BrainRegion.CDR',x,'.',unique(METADATA$CDR)),2), 2, paste, collapse = '-')
}) %>%
  as.character(), 
levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT = eBayes(FIT)

# Get differnetial expression
DE = list()
p = list()
for (i in 1:dim(contrast)[2]){
  DE[[i]] = topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id")
  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
  p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.CDR','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
}
# multiplot(plotlist=p, cols = 4)
names(DE) = gsub('BrainRegion.CDR','',colnames(contrast))

DE = DE %>% ldply(.id = 'Comparison')

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(Comparison, ensembl_gene_id) %>%
  group_by(Comparison) %>%
  dplyr::summarise(nDiffExp = length(unique(ensembl_gene_id)),
                   DiffExpGenes = paste(unique(ensembl_gene_id), collapse = ','))

kable(tmp[,1:2])
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "MSBB RNASeq (based on CDR)", parentId = parentId)
CODE <- synStore(CODE)

>>>>

---
title: "Covariates analysis of MSBB reprocessed RNASeq data"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)
library(githubr)

@@

knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_cdr_Reprocessed.Rmd",
                                 parentId ="syn5570248",
                                 entityName = 'MSBB Reprocessed RNASeq (based on CDR)')
```

@@

library(ggplot2)
library(reshape2)
library(limma)
library(RColorBrewer)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(xlsx)
library(ggdendro)

@@

cl = makeCluster(2)
registerDoParallel(cl)

devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

@@

thisFileName <- 'MSSM_RNASeq_cdr_Reprocessed.Rmd'

@@

Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download expression data
COUNT_ID <- 'syn6114414'
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, quote = '', header = T, sep = '\t')

# Get technical metadata
METADATA_ID <- 'syn5898489'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA_TECHNICAL <- read.table(synGet(METADATA_ID)@filePath,sep='\t',header=T, row.names=1) %>%
  rownameToFirstColumn('OldBarcode')

# Get clinical metadata
METADATA_CLINICAL_ID <- 'syn5475828'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_CLINICAL_ID
METADATA_CLINICAL <- read.xlsx(synGet(METADATA_CLINICAL_ID)@filePath, sheetIndex = 1)

# Manually fix 4 sample IDs
METADATA_CLINICAL[which(METADATA_CLINICAL$BB == 1132), "BM36.Fnl"] = 397
METADATA_CLINICAL[which(METADATA_CLINICAL$BB == 1117), "BM22.Fnl"] = 144
METADATA_CLINICAL[which(METADATA_CLINICAL$BB == 1320), "BM22.Fnl"] = 159
METADATA_CLINICAL[which(METADATA_CLINICAL$BB == 1475), "BM22.Fnl"] = 186

METADATA = left_join(METADATA_TECHNICAL %>%
                       dplyr::mutate(ID = gsub('hB_RNA_','',NewBarcode),
                                     ID = as.numeric(ID)),
                     METADATA_CLINICAL %>%
                       dplyr::select(BB, BM.44.Barcode, BM36.Fnl, BM22.Fnl, BM10.Fnl, PMI, RACE, 
                                     AOD, CDR, SEX, NP.1, PlaqueMean, bbscore) %>%
                       dplyr::rename(BM_44 = BM.44.Barcode,
                                     BM_10 = BM10.Fnl,
                                     BM_22 = BM22.Fnl,
                                     BM_36 = BM36.Fnl) %>%
                       tidyr::gather(BrainRegion, ID, -BB, -PMI, -RACE, 
                                     -AOD, -CDR, -SEX, -NP.1, -PlaqueMean, -bbscore) %>%
                       filter(!is.na(ID)) %>%
                       dplyr::mutate(ID = as.numeric(ID)))
rownames(METADATA) = METADATA$OldBarcode

# Filter data
ind <- intersect(rownames(METADATA), colnames(COUNT))
# ind <- setdiff(ind, 'hB_RNA_10622')
METADATA <- METADATA[ind,]
COUNT <- COUNT[, ind]
```
### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering, cache=TRUE}
# Max out age at 90
METADATA$AOD = gsub('\\+','',as.character(METADATA$AOD))

@@

# Select covariates of interest
FactorCovariates <- c('RACE', 'SEX', 'BrainRegion.CDR', 'Batch', 'BB')
ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2')

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
rownames(COVARIATES) <- METADATA$OldBarcode

@@

```{r dataExplore, fig.width = 10, fig.height = 10}
# RIN
p = list()
p[[1]] = ggplot(METADATA, aes(x = as.factor(CDR), y = as.numeric(RIN), fill = as.factor(BrainRegion))) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')

# AgeAtDeath
p[[2]] = ggplot(METADATA, aes(x = as.factor(CDR), y = as.numeric(AOD), fill = as.factor(BrainRegion))) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + theme(legend.position = 'top')

# Mean Plaque
p[[3]] = ggplot(METADATA, aes(x = as.factor(CDR), y = as.numeric(PlaqueMean), fill = as.factor(BrainRegion))) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PlaqueMean') + theme(legend.position = 'top')

# NP.1
p[[4]] = ggplot(METADATA, aes(x = as.factor(CDR), y = as.numeric(NP.1), fill = as.factor(BrainRegion))) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('NP.1') + theme(legend.position = 'top')

@@

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,FactorCovariates]))
```
```{r temp, include=F}
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, COVARIATES,'NULL design(voom-normalized)', isKeyPlot=TRUE, MIN_PVE_PCT_PC = 1)
```
Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

@@

backgroundGenes = data.frame(Gene.ID = rownames(RESIDUAL.GENE_EXPRESSION),
                             ID = rownames(RESIDUAL.GENE_EXPRESSION)) %>%
  tidyr::separate(ID, c('ensembl_gene_id','position'), sep = '\\.')

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "Gene.ID") %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>% 
  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, "Gene.ID") %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>% 
  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = as.data.frame(summary(COVARIATES$BrainRegion.CDR))
colnames(tmp) = 'No. of samples'
kable(tmp)
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=12, fig.width=25}
# Fit contrast
contrast = makeContrasts(contrasts = sapply(unique(METADATA$BrainRegion), function(x){
  apply(combn(paste0('BrainRegion.CDR',x,'.',unique(METADATA$CDR)),2), 2, paste, collapse = '-')
}) %>%
  as.character(), 
levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT = eBayes(FIT)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC, backgroundGenes){
  topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1]) %>%
    rownameToFirstColumn("Gene.ID") %>%
    left_join(backgroundGenes) %>%
    left_join(Ensemble2HGNC)
}, FIT, VOOM.ADJUSTED.GENE_EXPRESSION, Ensemble2HGNC, backgroundGenes)
names(DE) = gsub('BrainRegion.CDR','',colnames(contrast))

DE = DE %>% 
  ldply(.id = 'Comparison') %>%
  dplyr::mutate(Comparison1 = Comparison) %>%
  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
  tidyr::unite(Name, reference, against, sep = '-')

p <- ggplot(DE, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
p <- p + facet_grid(BrainRegion~.+Name)
p

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(BrainRegion, Name, ensembl_gene_id) %>%
  group_by(BrainRegion, Name) %>%
  dplyr::summarise(nDiffExp = length(unique(ensembl_gene_id))) %>%
  tidyr::spread(Name, nDiffExp)

kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "MSBB Reprocessed RNASeq (based on CDR)", parentId = parentId)
CODE <- synStore(CODE)
